---
title: "Coral Competition_Porites"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=FALSE, warning=FALSE, message=FALSE)
knitr::opts_chunk$set(cache=TRUE)
options(width = 60)

library(knitr)
library(DESeq2)
library(stringr)
library(xlsx)
library(tidyverse)
library(ggrepel)
library(pheatmap)
```

First we prepare the data.  This involves reading in a counts matrix from corset.  Then we extract the sample names from the columns and match those to conditions in an excel file that describes the experimental conditions for each sample.

```{r prepdata, include=FALSE}

# Obtained with `corset_alldifferent_genomics1.03.sh`
# First step to explore data: Using all samples
counts <- read_tsv("raw_data/03-counts.txt") %>% column_to_rownames(var="X1") %>% as.matrix()

sample_data <- read.csv("raw_data/Samples_data.csv",strip.white = T)
names(sample_data)

colnames(sample_data) <- c("ID","Tank","HardCoral",'treat',"SoftCoralControl","PdvsL_Other")
head(sample_data)

sample_data_row_order <- match(colnames(counts),sample_data$ID)

#Now we use DESeq2 to run the differential expression analyses
# This sets up the DESeq model
dds <- DESeqDataSetFromMatrix(countData = counts,colData=sample_data[sample_data_row_order,],design = ~  treat+ HardCoral)

# Now we run the actual DESeq fitting process. 
# We do this twice. The first is a standard fit that will allow wald tests between specific conditions. 
# The second is a likelihood ratio test (LRT) fit where we also fit a NULL model (without HardCoral) and compare it to the full model
##NOTE: is there a way to save dds ans ddslrt results?
if ( file.exists("data/dds.rds")){
  dds <- read_rds("data/dds.rds")
} else {
  dds <- DESeq(dds)
  write_rds(dds,"data/dds.rds")
}

if ( file.exists("data/ddslrt.rds")){
  ddslrt <- read_rds("data/ddslrt.rds")
} else {
  ddslrt <- DESeq(dds,test = "LRT",reduced = ~ 1 + HardCoral)
  write_rds(ddslrt,"data/ddslrt.rds")
}

# For each of the fits we extract results tables. 
res_dds<- results(dds)
resOrdered_dds <- as.data.frame(res_dds[order(res_dds$padj),]) %>% add_rownames("cluster_id") %>% filter(padj<0.1) # same as contrast Pd, Pf

# The LRT there is just one results table because it is a global test across all conditions
res_lrt <- results(ddslrt)

# For the standard fit we extract specific comparisons against control and treatment
res_T <- results(dds,contrast = c("treat","T","C"))
#Table DE genes when comparing Treatment and Control
resOrdered_T<- as.data.frame(res_T[order(res_T$padj),]) %>% add_rownames("cluster_id")%>% filter(padj<0.1)

# For the standard fit we extract specific comparisons against Pd and Pf
res_P <- results(dds,contrast = c("HardCoral","Pd","Pf"))
#Table DE genes when comparing Pd and Pf
resOrdered_P <- as.data.frame(res_P[order(res_P$padj),]) %>% add_rownames("cluster_id")%>% filter(padj<0.1)

####### Principal componet analysis #########

vsd <- varianceStabilizingTransformation(dds, blind=FALSE)
pca_data <- plotPCA(vsd,intgroup=c("HardCoral","SoftCoralControl"),returnData=TRUE)
  ggplot(pca_data,aes(x=PC1,y=PC2)) + geom_point(aes(color=SoftCoralControl,shape=HardCoral),size=4) + geom_label_repel(aes(label=group)) + xlab("xlabel") + ylab("customylabel") + theme(text = element_text(size=20))
```

# 2. Without PdLd as outlier, same model
```{r without PdLd}
counts2 <- data.frame(counts) %>% select(-PdLd) %>% as.matrix()

sample_data2 <- sample_data[match(colnames(counts2),sample_data$ID),]


# This sets up the DESeq model
dds2 <- DESeqDataSetFromMatrix(countData = counts2,colData=sample_data2,design = ~  treat+ HardCoral)

# Now we run the actual DESeq fitting process. 
dds2 <- DESeq(dds2)

ddslrt2 <- DESeq(dds2,test = "LRT",reduced = ~ 1 + HardCoral)

# For each of the fits we extract results tables. 
res_dds2<- results(dds2)
resOrdered_dds2 <- as.data.frame(res_dds2[order(res_dds2$padj),]) %>% add_rownames("cluster_id") %>% filter(padj<0.1) # same as contrast Pd, Pf

# The LRT there is just one results table because it is a global test across all conditions
res_lrt2 <- results(ddslrt2)

# For the standard fit we extract specific comparisons against control and treatment
res_T2 <- results(dds2,contrast = c("treat","T","C"))
#Table DE genes when comparing Treatment and Control
resOrdered_T2<- as.data.frame(res_T2[order(res_T2$padj),]) %>% add_rownames("cluster_id")%>% filter(padj<0.1)

# For the standard fit we extract specific comparisons against Pd and Pf
res_P2 <- results(dds2,contrast = c("HardCoral","Pd","Pf"))
#Table DE genes when comparing Pd and Pf
resOrdered_P2 <- as.data.frame(res_P2[order(res_P2$padj),]) %>% add_rownames("cluster_id")%>% filter(padj<0.1)

####### Principal componet analysis #########

vsd2 <- varianceStabilizingTransformation(dds2, blind=FALSE)
pca_data <- plotPCA(vsd2,intgroup=c("HardCoral","SoftCoralControl"),returnData=TRUE)
  ggplot(pca_data,aes(x=PC1,y=PC2)) + geom_point(aes(color=SoftCoralControl,shape=HardCoral),size=4) + geom_label_repel(aes(label=group))  + theme(text = element_text(size=20))
```
# 3. Changing model with SoftCoralControl with all samples
```{r, model : SoftCoralControl}

# This sets up the DESeq model
dds3 <- DESeqDataSetFromMatrix(countData = counts,colData=sample_data,design = ~  HardCoral+ SoftCoralControl)

# Now we run the actual DESeq fitting process. 
dds3 <- DESeq(dds3)

ddslrt3 <- DESeq(dds3,test = "LRT",reduced = ~ 1 + HardCoral)

res_dds3<- results(dds3)
resOrdered_dds3 <- as.data.frame(res_dds3[order(res_dds3$padj),]) %>% add_rownames("cluster_id") %>% filter(padj<0.1) #

# For the standard fit we extract specific comparisons against Pd and Pf
res_P3 <- results(dds3,contrast = c("HardCoral","Pd","Pf"))
#Table DE genes when comparing Pd and Pf
resOrdered_P3 <- as.data.frame(res_P3[order(res_P3$padj),]) %>% add_rownames("cluster_id")%>% filter(padj<0.1)

####### Principal componet analysis #########

vsd3 <- varianceStabilizingTransformation(dds3, blind=FALSE)
pca_data <- plotPCA(vsd3,intgroup=c("HardCoral","SoftCoralControl"),returnData=TRUE)
  ggplot(pca_data,aes(x=PC1,y=PC2)) + geom_point(aes(color=SoftCoralControl,shape=HardCoral),size=4) + geom_label_repel(aes(label=group)) + xlab("xlabel") + ylab("customylabel") + theme(text = element_text(size=20))
```

#4. Changing model with SoftCoralControl without PdLd

```{r, model : SoftCoralControl}

# This sets up the DESeq model
dds4 <- DESeqDataSetFromMatrix(countData = counts2,colData=sample_data2,design = ~  HardCoral+ SoftCoralControl)

# Now we run the actual DESeq fitting process. 
dds4 <- DESeq(dds4)

ddslrt4 <- DESeq(dds4,test = "LRT",reduced = ~ 1 + HardCoral)

res_dds4<- results(dds4)
resOrdered_dds4 <- as.data.frame(res_dds4[order(res_dds3$padj),]) %>% add_rownames("cluster_id") %>% filter(padj<0.1) #

# For the standard fit we extract specific comparisons against Pd and Pf
res_P4 <- results(dds4,contrast = c("HardCoral","Pd","Pf"))
#Table DE genes when comparing Pd and Pf
resOrdered_P4 <- as.data.frame(res_P4[order(res_P4$padj),]) %>% add_rownames("cluster_id")%>% filter(padj<0.1)
####### Principal componet analysis #########

vsd4 <- varianceStabilizingTransformation(dds4, blind=FALSE)
pca_data <- plotPCA(vsd4,intgroup=c("HardCoral","SoftCoralControl"),returnData=TRUE)
  ggplot(pca_data,aes(x=PC1,y=PC2)) + geom_point(aes(color=SoftCoralControl,shape=HardCoral),size=4) + geom_label_repel(aes(label=group)) + theme(text = element_text(size=20))
```

#5. Changing model with SoftCoralControl + PdvsL_Other without PdLd

```{r, model : SoftCoralControl}

# This sets up the DESeq model
dds5 <- DESeqDataSetFromMatrix(countData = counts2,colData=sample_data2,design = ~  HardCoral+ SoftCoralControl + PdvsL_Other)

# Now we run the actual DESeq fitting process. 
dds5 <- DESeq(dds5)

ddslrt5 <- DESeq(dds5,test = "LRT",reduced = ~ 1 + HardCoral)
res_dds5<- results(dds5)
resOrdered_dds5<- as.data.frame(res_dds5[order(res_dds4$padj),]) %>% add_rownames("cluster_id") %>% filter(padj<0.1) #

# For the standard fit we extract specific comparisons against Pd and Pf
res_P5 <- results(dds5,contrast = c("HardCoral","Pd","Pf"))
#Table DE genes when comparing Pd and Pf
resOrdered_P5 <- as.data.frame(res_P5[order(res_P5$padj),]) %>% add_rownames("cluster_id")%>% filter(padj<0.1)

# For the standard fit we extract specific comparisons against Pd and Pf
res_O <- results(dds5,contrast = c("PdvsL_Other","Pd","O"))
#Table DE genes when comparing Pd and Pf
resOrdered_O <- as.data.frame(res_O[order(res_O$padj),]) %>% add_rownames("cluster_id")%>% filter(padj<0.1)


####### Principal componet analysis #########

vsd5 <- varianceStabilizingTransformation(dds4, blind=FALSE)
pca_data <- plotPCA(vsd5,intgroup=c("HardCoral","SoftCoralControl"),returnData=TRUE)
  ggplot(pca_data,aes(x=PC1,y=PC2)) + geom_point(aes(color=SoftCoralControl,shape=HardCoral),size=4) + geom_label_repel(aes(label=group)) + theme(text = element_text(size=20))
```
