---
title: "DESeq analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=FALSE, warning=FALSE, message=FALSE)
knitr::opts_chunk$set(cache=TRUE)
options(width = 60)

library(knitr)
library(DESeq2)
library(tidyverse)
library(ggrepel)
library(pheatmap)
```



Counts obtained from Corset were analysed with DESeq to identify differentially expressed genes between HardCoral treatments. 

Initial data exploration with PCA revealed that PdLd is an extreme outlier.  We therefore excluded this sample from further analysis


```{r prepdata, include=FALSE}
#First we prepare the data.  
#This involves reading in a counts matrix from corset.  
#Then we extract the sample names from the columns and match those to conditions in an excel file that describes the experimental conditions for each sample.

counts <- read_tsv("hpc/corset/03-counts.txt") %>% 
  column_to_rownames(var="X1") %>% 
  as.matrix()

sample_data <- read_csv("raw_data/Samples_data.csv",trim_ws = TRUE)
colnames(sample_data) <- c("ID","Tank","HardCoral",'treat',"SoftCoralControl","PdvsL_Other")

sample_data_row_order <- match(colnames(counts),sample_data$ID)

dds <- DESeqDataSetFromMatrix(countData = counts,colData=sample_data[sample_data_row_order,],design = ~  treat+ HardCoral)
vsd <- varianceStabilizingTransformation(dds, blind=FALSE)

plotPCA(vsd,intgroup=c("HardCoral","SoftCoralControl"),returnData=TRUE) %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(color=SoftCoralControl,shape=HardCoral),size=4) + 
  geom_label_repel(aes(label=group)) + 
  xlab("xlabel") + 
  ylab("customylabel") + 
  theme(text = element_text(size=20))
```


Differential Expression Analysis with the outlier excluded was then performed based on a model including the treatment, HardCoral and their interaction

> ~ treat*HardCoral

```{r}
# 2. I eliminate PdLd from the analysis and refit the same model : treat + HardCoral
# 2.1 Results for contrast : treat, "T", "C" > just 58 genes DE
# 2.2 Results for contrasr: HardCoral, "Pd", "Pf"> 37661 genes DE btw colonies which is not a very intresting result
# 2.2 PCA for dds > shows a difference btw colonies as well but it shows differences among the different nubbins from the same colony

counts2 <- data.frame(counts) %>% select(-PdLd) %>% as.matrix()

sample_data2 <- sample_data[match(colnames(counts2),sample_data$ID),]

dds2 <- DESeqDataSetFromMatrix(countData = counts2,colData=sample_data2,design = ~  treat*HardCoral)

if (file.exists("cache/dds2.rds")) {
    dds2 <- read_rds("cache/dds2.rds")
} else {
  dds2 <- DESeq(dds2)
  write_rds(dds2,"cache/dds2.rds")
}

resultsNames(dds2)
```


```{r}

# This contrast gives genes consistently DE between treatment and control
res_T2 <- results(dds2,contrast = c("treat","T","C"),tidy = TRUE) %>% 
  rename(cluster_id=row) %>% 
  arrange(padj) %>% 
  filter(padj<0.05)
  
# This contrast looks for genes where the treatment and HardCoral interact
res_I2 <- results(dds2,name="treatT.HardCoralPf",tidy = TRUE) %>% 
  rename(cluster_id=row) %>% 
  arrange(padj) %>% 
  filter(padj<0.05)

# This just looks for genes DE between HardCorals.  There are so many of these it is meaningless
res_P2 <- results(dds2,contrast = c("HardCoral","Pd","Pf"),tidy = TRUE) %>% 
  rename(cluster_id=row) %>% 
  arrange(padj) %>% 
  filter(padj<0.05)

vsd2 <- varianceStabilizingTransformation(dds2, blind=FALSE)

plotPCA(vsd2,intgroup=c("HardCoral","SoftCoralControl"),returnData=TRUE) %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(color=SoftCoralControl,shape=HardCoral),size=4) + 
  geom_label_repel(aes(label=group))  + 
  theme(text = element_text(size=20))
```

For genes differentially expressed between control and treatment we plot raw values as scatterplots

```{r}
# For genes DE ONLY in  contrast btw Control and treatment

t2_genes_tidy <- assay(vsd2)[which(rownames(vsd2) %in% res_T2$cluster_id),] %>% 
  as.data.frame() %>% 
  rownames_to_column("cluster_id") 

t2_genes_tidy %>% 
  pivot_longer(-cluster_id,names_to = "ID",values_to="count") %>% 
  left_join(sample_data) %>% 
  ggplot(aes(x=SoftCoralControl, y=log(count), colour=HardCoral)) + 
  geom_point(aes(shape=treat)) + 
  facet_wrap(~cluster_id,ncol = 4)
```

Now plot raw data as heatmaps.  For this the relative change in expression is plotted (relative to the mean for a gene) so that clustering is meaningful


```{r}
t2_heatmap_data <- t2_genes_tidy %>% 
  as.data.frame() %>% 
  column_to_rownames("cluster_id")

t2_heatmap_data_relative <- sweep(t2_heatmap_data, MARGIN=1, STATS= rowMeans(t2_heatmap_data))

# This writes the heatmap to a file.  
# In this case the file is deliberately much larger than one page so all the proteins can be displayed with their names.
# If you want a pdf to include in a talk the names will probably need to be removed (show_rownames=FALSE) and the size adjusted accordingly
#
#pdf(file="results/heatmap_58.pdf",height =20,width = 20)
pheatmap(t2_heatmap_data_relative, cluster_rows=TRUE, show_rownames=TRUE,cluster_cols=FALSE)
#dev.off()  
  
```

Plot a heatmap for genes with significant interactions.  This is interesting because it seems to indicate that there are many genes where Pf upregulates the gene in interaction relative to control whereas Pd does not.


```{r}
# For genes DE ONLY in  contrast btw Control and treatment

i2_genes_tidy <- assay(vsd2)[which(rownames(vsd2) %in% res_I2$cluster_id),] %>% 
  as.data.frame() %>% 
  rownames_to_column("cluster_id") 

i2_heatmap_data <- i2_genes_tidy %>% 
  as.data.frame() %>% 
  column_to_rownames("cluster_id")

i2_heatmap_data_relative <- sweep(i2_heatmap_data, MARGIN=1, STATS= rowMeans(i2_heatmap_data))

pheatmap(i2_heatmap_data_relative, cluster_rows=TRUE, show_rownames=TRUE,cluster_cols=FALSE)
```