---
title: "Polyp Activity. Plots and Final Statistics"
author: "Rhondda Jones, Natalia Andrade and Ira Cooke"
date: "29/08/2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ordinal)
library(cowplot)
library(grid)
library(ggrepel)
library(DESeq2)
library(ggpubr)
```

```{r}
polyp_act<-read_csv("raw_data/Porites_polypact.csv") %>% 
  mutate_if(is.numeric, as.factor)

polyp_act_long <- polyp_act %>% 
  pivot_longer(cols = starts_with("DAY"), names_to="Time", values_to="Activity") %>% 
  mutate(Time=as.integer(str_match(Time, pattern = "DAY([0-9]+)")[,2])) %>% 
  mutate(Activity=factor(Activity, levels = c("C", "P","O"), ordered=TRUE)) %>% 
  mutate(TANK=as.factor(TANK)) %>% 
  mutate(TimeCat = cut(Time, breaks=c(0,14.1,19.1,24.1,29.1,34.1,39.1,44.1,49.1,54.1,60))) %>% 
  mutate(TimeCat = factor(TimeCat,
         labels=c("<15","15-20","20-25","25-30","30-35","35-40","40-45","45-50","50-55","55-60")))
```

Summarise the distribution of replicates across experimental factors.

```{r}
polyp_act_long %>% 
  group_by(TimeCat,TANK,HARD,TREAT,SOFT) %>% 
  summarise(count=n()) %>% 
  ggplot(aes(x=TimeCat)) + 
  geom_tile(aes(fill=as.factor(count),y=TANK)) +
  geom_point(aes(y=TANK,color=SOFT),size=0.5) +
  facet_wrap(~HARD*TREAT,nrow = 1) + theme(axis.text.x = element_text(angle=90)) + 
  scale_fill_grey()
```


This is a ggplot equivalent of of the Mosaicplots used in data exploration. This shows data aggregated across nubbins and tanks which does hide some important variation that is hard to capture graphically (see stats).  Overall however there are two clear trends;

1. The treated colonies are more likely to be partial or closed than controls
2. Colonies seemed to be more likely to be partial or closed at the start of the trial and gradually converted to open as the trial progressed.

```{r}
polyp_act_long %>% na.omit() %>% 
  group_by(HARD,TREAT,TimeCat) %>%   
  mutate(tankcount=length(unique(TANK))) %>% 
  group_by(HARD,TREAT,TimeCat,Activity) %>% 
  mutate(catcount=n()) %>% 
  ungroup() %>% 
  mutate(TREAT = ifelse(TREAT=="Yes","Competition","Control")) %>% 
  mutate(TimeCat = as.factor(TimeCat)) %>% 
  ggplot(aes(x=TimeCat)) + 
  geom_col(aes(fill=Activity,y=catcount), position = "fill") +
  geom_text(aes(y=0.1,label=tankcount), fill = "transparent",size=2, color="grey40") +
  geom_vline(xintercept=4.5,color="blue") +
  facet_grid(HARD~TREAT) + 
  theme_pubclean() +
  theme(axis.text.x = element_text(angle=90)) + 
  xlab("Days Since Start") + ylab("Proportion of Nubbins")
```


```{r, echo=FALSE, eval=FALSE}
ggsave("figures/polyp_activity.png",width = 12,height = 8)
```


Based on extensive model selection analysis (see [03_polyp_activity_exploration.md](03_polyp_activity_exploration.md)) a generalised cumulative mixed effects model was chosen to fit the polyp activity data. 


```{r}
fm3d =clmm2(location=Activity ~ TimeCat+TREAT+HARD, random=TANK, Hess=TRUE,  data=polyp_act_long)
summary(fm3d)
```

## Mesenteric Filament Data

```{r}
mesenteries_data <- readxl::read_excel("~/Dropbox/natalia_competition/Porites_competition/Mesentries Obsv_Bothcorals.xlsx") %>%
  dplyr::rename(HARD=`Hard corasls`) %>% 
  filter(HARD %in% c("Pa","Pc")) %>% 
  mutate(HARD = ifelse(HARD=="Pa","Pd","Pf"))

mesenteries_groups <- mesenteries_data %>% unite("pair",HARD,Softies,sep="") %>% pull(pair) %>% unique()
```


